### 2020-12-04 ch01_introduction/introduction.tex
Finished section of ESMs

>>>>

\section{Motivation}
\label{ch:01:motivation}

- Why is climate change relevant for the society?

- Cite latest warming reports, also other variables (droughts, extreme events,
etc.)

- basic physics of greenhouse gases (vibration modes, etc.)

- well-mixed greenhouse gases (water vapor, CO2, CH4) + sources

- long-lived vs. short-lived GHGs?

- aerosols + cooling effect

- natural vs. anthropogenic effects (D \& A) -> clear evidence that climate
change over 21st century is caused by human influence
(https://globalwarmingindex.org/, Haustein et al., 2017)

### 2020-12-04 ch02_scientific_background/scientific_background.tex
Finished section of ESMs

- anthropogenic forcing vs. natural forcing (volcanoes, solar cycle, etc.)

- basic physics of greenhouse gases (vibration modes, etc.)

- historical vs. future: in historical (future), uncertainty in forcing low
(high)

- definition of feedbacks, examples

\section{Climate Sensitivity}
\label{sec:02:climate_sensitivity}

@@

Thus, this removal of \ac{CO2} from the atmosphere actively slows down global
warming. However, whether this benefit will persist in the future remains
unclear, which is primarily linked to two feedback processes connecting the
climate system and the global carbon cycle: the \emph{concentration-carbon
  feedback} and the \emph{climate-carbon feedback}
\autocite{Friedlingstein2006, Gregory2009, Collins2013}. For the terrestrial
biosphere, the concentration-carbon feedback is connected to the \emph{\ac{CO2}
  fertilization effect} \autocite{Walker2020}, that causes an increase of
photosynthesis rates when the atmospheric \ac{CO2} concentration increases,
which in turns removes \ac{CO2} from the atmosphere, forming a negative
feedback. For the ocean, the concentration-carbon feedback is negative as well.
In this case, an elevated atmospheric \ac{CO2} concentration causes an
increased dissolution of \ac{CO2} into the sea, which increases the ocean
carbon uptake. On the other hand, the climate-carbon feedback is thought to be
positive for both the terrestrial biosphere and the ocean
\autocite{Gregory2009}. In the first case, temperature and precipitation
changes due to anthropogenic activities decrease the land carbon uptake because
of increased temperature and water stress on photosynthesis and higher
ecosystem respiration costs, which accelerates global warming due to more
\ac{CO2} that remains in the atmosphere. For the ocean, increased temperatures
lead to a reduction of vertical transport in the ocean resulting from increased
stability and reduced solubility of \ac{CO2} in the sea, which reduces the
ocean carbon uptake and enhances climate change \autocite{Gregory2009}.

@@

>>>>

Simulations from climate model ensembles of \ac{CMIP} allow us to assess the
impact of future climate change in a consistent and transparent way. Especially
the \ac{ScenarioMIP} experiments can give valuable insights into the upcoming
development of the Earth system by providing \emph{projections} of the future
climate. In contrast to climate predictions, climate projections run over
multiple decades and depend upon the future scenario considered, which are
based on assumptions that may or may not turn out to be correct. On the
contrary, climate predictions are attempts to predict the actual evolution of
the climate on much shorter time scales from seasons to years. Similar to any
other scientific experiment, climate model projections suffer from associated
uncertainties. There are three major sources of climate model projections we
can distinguish: natural variability, climate response uncertainty and emission
uncertainty \autocite{Hawkins2009, Hawkins2010}.
\Cref{fig:02:sources_of_uncertainty} shows these three sources for the
projected global mean surface temperature anomaly over the \nth{21} century.

\begin{figure}[t]
  \centering
  \includegraphics[width=\LargeFigureWidth{}]{
    ch02_scientific_background/figs/sources_of_uncertainty.pdf}
  \caption{Schematic illustration of the relative importance of different
    sources of uncertainties in climate model projections and their evolution
    in time. (a) Time series of the anomaly of the decadal and global mean
    surface temperature relative to the period \range{1961}{1980}. The black
    line shows the historical observations with estimates of uncertainty from
    climate models (gray). The remaining colors show different sources of
    uncertainty in future climate projections: Natural variability
    (orange), climate response uncertainty (blue) and emission uncertainty
    (green) \autocite{Hawkins2009, Hawkins2010}. Climate response uncertainty
    can (b) increase in newer generations of climate models when a new process
    is discovered to be relevant or (c) decrease with additional model
    improvements and observational constraints. Taken from
    \textcite{Cubasch2013}.}
  \label{fig:02:sources_of_uncertainty}
\end{figure}

\emph{Natural variability} is connected to the chaotic nature of the Earth
system that arises from complex interactions between the ocean, atmosphere,
land, biosphere and cryosphere \autocite{Cubasch2013}. It constitutes a
fundamental limit of how precisely we can project the future climate since it
is inherent in the Earth system and cannot be eliminated by more knowledge and
more advanced climate models. Natural variability is more relevant on regional
and local scales than on continental or global scales. Further contributions to
natural variability on longer time scales come from phenomena like the
\ac{ENSO} or the \ac{NAO} and from externally (and thus explainable) events
like volcanic eruptions and variations in the solar activity. Natural
variability can be seen as the \emph{noise} in the climate record as opposed to
the anthropogenic \emph{signal} \autocite{Cubasch2013}. As illustrated by
\cref{fig:02:sources_of_uncertainty}, the uncertainty associated with natural
variability is constant over time.

The second source of uncertainty in climate model projections is \emph{emission
  uncertainty}. This arises from the different possible trajectories in terms
of future forcing (\acp{GHG}, aerosols, land use changes, \etc{}) humankind
might take. Examples for these are the \ac{SSP}-based experiments given by
\ac{ScenarioMIP} that include a variety of different scenarios from a
sustainable future to a full fossil fuel-based development (see
\cref{subsec:02:cmip}). A possible approach to quantify emission uncertainty is
to assess the climate impact of these different trajectories. Since the
emission uncertainty strongly depends on the future development of the human
society, it cannot be reduced by improving climate models. In contrast to
natural variability, the emission uncertainty increases over time in climate
projections, since estimating forcings for the near future is easier than for
the far future.

Finally, the third source of uncertainty in climate model projections is the
\emph{climate response uncertainty}, which comes from our imperfect knowledge
of how the climate system will respond to future anthropogenic forcings. Due to
the complexity of the Earth system, the future climate could develop in many
different ways that are all consistent with our current knowledge and models
\autocite{Cubasch2013}. In the context of climate model ensembles, the climate
response uncertainty is often also called \emph{model uncertainty} and reflects
the different responses of the different climate models to a given forcing.
Even though all climate models are built on the same physical principles, they
differ in terms of spatial resolution, processes included and parametrizations
of unresolved processes. The latter can be thought of as different
approximations that are necessary to represent processes that take place on
scales smaller than the common size of grid boxes in modern global climate
models (about $\ang{1} \times \ang{1}$ horizontally), which is for example the
case for some cloud processes and vegetation processes.

These differences in the climate models also give rise to different intensities
of \emph{climate feedbacks} (or even their presence/absence) in the models. A
climate feedback is a mechanism that either amplifies (\emph{positive
  feedback}) or weakens (\emph{negative feedback}) the effect of an external
forcing. An example of a positive feedback is the water vapor feedback, where
the increased surface temperature (caused by anthropogenic forcing) leads to an
enhanced evaporation of water which enhances the amount of water vapor in the
atmosphere. Since water vapor itself is a powerful \ac{GHG}, this amplifies the
effect of the anthropogenic forcing by further increasing the surface
temperatures \autocite{Cubasch2013}. Further examples and a mathematical
description of feedbacks are given in
\cref{sec:02:feedbacks_and_climate_sensitivity}.

As sciences evolves, representations of already included processes can be
improved in climate models. Moreover, new geophysical and biogeochemical
processes can be added to them. On the one hand, this can increase the climate
response uncertainty when a new process is discovered to be relevant
\commentcite{Cubasch2013}{see \cref{fig:02:sources_of_uncertainty}b}. However,
such an increase corresponds to a previously unmeasured uncertainty. An example
for this has recently happened in \acs{CMIP}6: most likely due to changes in
the cloud representation of the models the spread in the projected global mean
surface air temperature caused by a doubling of the atmospheric \ac{CO2}
concentration has substantially increased in \ac{CMIP}6 compared to older
\ac{CMIP} generations \autocite{Zelinka2020}. On the other hand, the climate
response uncertainty can decrease with additional model improvements and better
understanding of the Earth system (see \cref{fig:02:sources_of_uncertainty}c).
Moreover, it can also be reduced by observational constraints, which is the
main topic of this thesis.

\section{Feedback Processes and Climate Sensitivity}
\label{sec:02:feedbacks_and_climate_sensitivity}

- Overview over feedbacks (Cubasch et al., 2013, Figure 1.2)

- Mathematical description of feedbacks, geometric series

@@

Thus, this removal of \ac{CO2} from the atmosphere actively slows down global
warming. However, whether this benefit will persist in the future remains
unclear, which is primarily linked to two feedback processes connecting the
physical climate system and the global carbon cycle: the
\emph{concentration-carbon feedback} and the \emph{climate-carbon feedback}
\autocite{Friedlingstein2006, Gregory2009, Collins2013}. For the terrestrial
biosphere, the concentration-carbon feedback is connected to the \emph{\ac{CO2}
  fertilization effect} \autocite{Walker2020}, that causes an increase of
photosynthesis rates when the atmospheric \ac{CO2} concentration increases,
which in turns removes \ac{CO2} from the atmosphere, forming a negative
feedback. For the ocean, the concentration-carbon feedback is negative as well.
In this case, an elevated atmospheric \ac{CO2} concentration causes an
increased dissolution of \ac{CO2} into the sea, which increases the ocean
carbon uptake. On the other hand, the climate-carbon feedback is thought to be
positive for both the terrestrial biosphere and the ocean
\autocite{Gregory2009}. In the first case, temperature and precipitation
changes due to anthropogenic activities decrease the land carbon uptake because
of increased temperature and water stress on photosynthesis and higher
ecosystem respiration costs, which accelerates global warming due to more
\ac{CO2} that remains in the atmosphere. For the ocean, increased temperatures
lead to a reduction of vertical transport in the ocean resulting from increased
stability and reduced solubility of \ac{CO2} in the sea, which reduces the
ocean carbon uptake and enhances climate change \autocite{Gregory2009}.

@@

- Problems with multi-model analysis (search for refs in Schlund et al., ESD,
2020) -> models are not independent, multi-model ensemble is not "real"
statistical sample
